---
title: "Reproducible Research: Peer Assessment 1"
author: Victor Zambrano
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r first_chunk}
stepsDay2 <- read.csv('data/activity.csv')
minuteAverage <- tapply(stepsDay2$steps, stepsDay2$interval, mean, na.rm = T)
minuteAverage <- as.data.frame(minuteAverage)
```

Now we process the data to substitute NA values with the average steps taken on a 5-minute time frame

```{r process_data}
stepsDay2$processed <- 0
for (iter in 1:length(stepsDay2$steps)){
  if (is.na(stepsDay2$steps)){
    stepsDay2$processed[iter] <- minuteAverage[[((iter-1) %% length(minuteAverage))+1]]
  } else {
    stepsDay2$processed[iter] <- stepsDay2$steps[iter]
  }
}
```


## What is mean total number of steps taken per day?

In order to calculate the mean over different days, we use de tapply function.

```{r second_chunk}
dailySteps <- tapply(stepsDay2$steps, stepsDay2$date, sum)
stepsPerDay <- mean(dailySteps)
```

Each day the subject takes `stepsPerDay` steps per day on average. Over the 61 days that the
test was taken, the average steps taken per day is described in the following histogram

```{r histogram_chunk}
hist(dailySteps, main = 'Daily Steps Average', col = 'blue')
```



## What is the average daily activity pattern?


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
